---
title: ""
pagetitle: "Blue Wave squad report"
output:
  html_document:
    self_contained: true
    theme: null
    highlight: null
    mathjax: null
    css: blue-wave-report.css
params:
  team: "CUW-M"
  data_dir: "data"
---

```{r setup, include=FALSE}
# Iteration: 1
knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE,
  fig.width = 9, fig.height = 4.8, dpi = 140, dev = "png"
)
suppressPackageStartupMessages({
  library(tidyverse); library(knitr); library(broom)
})

ld <- list(blue = "#44759e", orange = "#f38439", beige = "#dee3c8",
           green = "#749c4c", gray = "#605b54", ink = "#1d2433")

theme_report <- function(base_size = 13) {
  theme_minimal(base_size = base_size, base_family = "Georgia") +
    theme(
      panel.grid.minor = element_blank(),
      plot.title    = element_text(face = "bold", size = base_size + 1),
      plot.subtitle = element_text(color = "#555555", margin = margin(b = 8)),
      plot.caption  = element_text(color = "#777777", size = base_size - 4, hjust = 0),
      axis.title    = element_text(color = "#444444"),
      strip.text    = element_text(face = "bold")
    )
}

tier_colours <- c(
  "European professional"     = ld$blue,
  "European amateur or lower" = ld$green,
  "Reserve or youth"          = ld$beige,
  "Island league"             = ld$orange,
  "Unknown"                   = ld$gray
)

pct <- function(x, d = 0) sprintf(paste0("%.", d, "f%%"), 100 * x)
```

```{r load}
squad <- read_csv(file.path(params$data_dir, "blue_wave_squad.csv"),
                  show_col_types = FALSE) |>
  mutate(
    island = if_else(str_starts(team_code, "CUW"), "Curaçao", "Aruba"),
    gender = if_else(str_ends(team_code, "M"), "Men", "Women"),
    home_code = if_else(str_starts(team_code, "CUW"), "CUW", "ABW"),
    based_abroad = !(club_country %in% c("CUW", "ABW", "X")),
    tier_group = case_when(
      league_tier %in% c("1", "2") ~ "European professional",
      league_tier == "3"           ~ "European amateur or lower",
      league_tier %in% c("R", "Y") ~ "Reserve or youth",
      league_tier == "L"           ~ "Island league",
      .default                     = "Unknown"
    )
  )

fifa <- read_csv(file.path(params$data_dir, "fifa_rankings.csv"),
                 show_col_types = FALSE)

this <- filter(squad, team_code == params$team)
this_island <- unique(this$island)
this_gender <- unique(this$gender)
this_iso    <- unique(this$home_code)

fifa_row <- filter(fifa, iso3 == this_iso)

n_players    <- nrow(this)
n_abroad     <- sum(this$based_abroad)
share_abroad <- n_abroad / n_players
n_topflight  <- sum(this$league_tier %in% c("1", "2"))
n_countries  <- n_distinct(this$club_country[this$club_country != "X"])
n_unknown    <- sum(this$club_country == "X")
n_home       <- sum(this$club_country == this_iso)
```

<div class="cover">
<div>

<div class="cover-tag">DCDC Network &middot; Blue Wave Analytics</div>

<div class="cover-title">`r this_island`</div>

<div class="cover-sub">`r this_gender`'s national squad, 2026: where `r n_players` players actually play their club football</div>

<div class="cover-accent"></div>

</div>

<div class="cover-stats">
  <div class="cs-item"><div class="cs-num">`r n_players`</div><div class="cs-lbl">Players</div></div>
  <div class="cs-item"><div class="cs-num">`r pct(share_abroad)`</div><div class="cs-lbl">Abroad</div></div>
  <div class="cs-item"><div class="cs-num">`r n_topflight`</div><div class="cs-lbl">Top two tiers</div></div>
  <div class="cs-item"><div class="cs-num">`r n_home`</div><div class="cs-lbl">Island league</div></div>
  <div class="cs-item"><div class="cs-num">`r n_countries`</div><div class="cs-lbl">Countries</div></div>
  <div class="cs-item"><div class="cs-num">`r if (nrow(fifa_row)) fifa_row$rank else NA`</div><div class="cs-lbl">FIFA rank</div></div>
</div>

<div class="cover-footer">
Produced from a single R Markdown file. Every number on this page is computed
from <strong>blue_wave_squad.csv</strong> at knit time. Change the data, press
Knit, and the report rewrites itself. <br>
<strong>Generated:</strong> `r format(Sys.Date(), "%d %B %Y")`
</div>
</div>

## The question

`r this_island` has a population of about
`r if (nrow(fifa_row) && !is.na(fifa_row$population)) format(round(fifa_row$population), big.mark = ",") else "unknown"`
people. Its `r tolower(this_gender)`'s national squad has `r n_players` players
in it. Of those, `r n_abroad` play their club football somewhere other than the
island, which is `r pct(share_abroad)` of the squad, spread across
`r n_countries` countries.

`r if (n_home == 0) sprintf("Not one player in this squad plays club football on %s.", this_island) else sprintf("%d of them play in the island league.", n_home)`

## Where they play

```{r map-countries, fig.height=4.2}
this |>
  filter(club_country != "X") |>
  count(club_country, sort = TRUE) |>
  mutate(club_country = fct_reorder(club_country, n)) |>
  ggplot(aes(x = club_country, y = n)) +
  geom_col(fill = ld$blue, width = 0.7) +
  coord_flip() +
  labs(
    title = sprintf("Club country, %s %s squad", this_island, tolower(this_gender)),
    subtitle = if (n_unknown > 0)
      sprintf("%d further players could not be placed and are excluded", n_unknown)
      else NULL,
    x = NULL, y = "Players",
    caption = "Source: Cornerstone Economics squad dataset, 2026"
  ) +
  theme_report()
```

## At what level

Country of club is only half the story. A player at a Dutch third-tier amateur
side and a player in the Eredivisie are both "abroad", and they are not the same
thing. Grouping by the level of the league they play in separates them.

```{r level, fig.height=3.6}
this |>
  count(tier_group) |>
  mutate(tier_group = fct_reorder(tier_group, n)) |>
  ggplot(aes(x = tier_group, y = n, fill = tier_group)) +
  geom_col(width = 0.7) +
  scale_fill_manual(values = tier_colours) +
  coord_flip() +
  labs(title = "Level of club football", x = NULL, y = "Players") +
  theme_report() +
  theme(legend.position = "none")
```

## Against the neighbours

The same table for all four ABC island squads puts this one in context.

```{r compare}
comparison <- squad |>
  group_by(island, gender) |>
  summarise(
    players     = n(),
    abroad      = sum(based_abroad),
    pct_abroad  = round(100 * mean(based_abroad)),
    top_two     = sum(league_tier %in% c("1", "2")),
    island_league = sum(league_tier == "L"),
    .groups = "drop"
  ) |>
  arrange(desc(pct_abroad))

kable(comparison, col.names = c("Island", "Squad", "Players", "Abroad",
                                "% abroad", "Top two tiers", "Island league"))
```

```{r compare-chart, fig.height=3.4}
squad |>
  count(team_code, tier_group) |>
  ggplot(aes(x = team_code, y = n, fill = tier_group)) +
  geom_col(position = "fill", width = 0.7) +
  scale_fill_manual(values = tier_colours) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Squad composition, all four ABC island teams",
    x = NULL, y = NULL, fill = NULL
  ) +
  theme_report(12) +
  theme(legend.position = "bottom")
```

```{r chisq-calc}
sq <- squad |>
  filter(club_country != "X") |>
  mutate(where_playing = if_else(club_country %in% c("CUW", "ABW"),
                                 "Island league", "Abroad"))
ct <- table(sq$island, sq$where_playing)
cs <- suppressWarnings(chisq.test(ct))
```

<div class="callout">
<div class="callout-h">Is the difference real?</div>
<p>A chi-square test of independence on the two islands' squads gives
X-squared = `r round(cs$statistic, 2)` on `r cs$parameter` degree of freedom,
p = `r format.pval(cs$p.value, digits = 2)`. The two islands build their squads
differently, and the gap is larger than sampling noise would comfortably
produce. `r n_unknown_all <- sum(squad$club_country == "X"); sprintf("%d of %d players were excluded from this test because their club could not be established.", n_unknown_all, nrow(squad))`</p>
</div>

## Punching above the population

```{r model}
reg <- lm(points ~ log_population + diaspora_per_capita, data = fifa)

fifa_resid <- fifa |>
  filter(!is.na(log_population), !is.na(diaspora_per_capita)) |>
  mutate(
    predicted = predict(reg, newdata = pick(everything())),
    residual  = points - predicted
  )

this_resid <- filter(fifa_resid, iso3 == this_iso)
```

A country's FIFA points are, unsurprisingly, related to how many people it has.
Regressing points on log population and diaspora size across
`r nrow(fifa_resid)` national associations explains about
`r pct(summary(reg)$r.squared)` of the variance. Population carries the model;
the diaspora term is not statistically distinguishable from zero.

What that model gives us is a benchmark. The residual, actual points minus
predicted points, says who over-performs for their size.

```{r resid-chart, fig.height=4.6}
focus <- c("CUW", "ABW", "JAM", "ISL", "MLT", "TTO", "SUR", "CPV")

fifa_resid |>
  mutate(highlight = iso3 %in% focus) |>
  ggplot(aes(x = log_population, y = points)) +
  geom_point(color = "grey80", size = 2) +
  geom_smooth(method = "lm", se = FALSE, color = ld$gray, linewidth = 0.6) +
  geom_point(data = ~ filter(.x, highlight), color = ld$orange, size = 3.4) +
  geom_text(
    data = ~ filter(.x, highlight),
    aes(label = country), nudge_y = 70, size = 3.3, family = "Georgia"
  ) +
  labs(
    title = "FIFA points against population",
    subtitle = "Countries above the line score more than their size predicts",
    x = "Population (log scale)", y = "FIFA points",
    caption = "Source: FIFA rankings; UN population estimates"
  ) +
  theme_report()
```

```{r resid-table}
fifa_resid |>
  filter(iso3 %in% focus) |>
  select(country, rank, points, predicted, residual) |>
  mutate(across(where(is.numeric), \(x) round(x, 0))) |>
  arrange(desc(residual)) |>
  kable(col.names = c("Country", "FIFA rank", "Points", "Predicted", "Residual"))
```

`r if (nrow(this_resid)) sprintf("%s scores %s points, against %s predicted for a country its size: a residual of %s.", this_island, round(this_resid$points), round(this_resid$predicted), round(this_resid$residual)) else ""`

## The squad

```{r squad-table}
this |>
  arrange(position, player_name) |>
  select(player_name, position, club, club_country, league_tier, confidence) |>
  kable(col.names = c("Player", "Position", "Club", "Country", "Tier", "Confidence"))
```

<div class="callout">
<div class="callout-h">What this dataset cannot tell you</div>
<p>The squad list was compiled from federation announcements, Wikipedia squad
tables, and commercial football databases, and each record carries a confidence
code. `r sum(this$confidence == "H")` of `r n_players` entries in this squad are
verified against a named recent source; the rest rest on a single source or an
inference. `r n_unknown` players could not be placed at a club at all.</p>
<p>Squad composition is also a snapshot of one call-up, not a measure of a
football system. It says nothing about youth development, coaching, or the
domestic game, and a report that reads it as such is reading too much into
`r n_players` rows.</p>
</div>

<div class="cover-footer">
Blue Wave Analytics, a DCDC Network training deliverable. Data and course
materials CC-BY 4.0. Squad dataset compiled for Cornerstone Economics working
paper WP-2026-01.
</div>
